{\Huge Supersymmetry} \\ \vspace{5mm} Lecture notes for FYS5190/FYS9190 Supersymmetry

### 2023-08-22 groups.tex
Updates to 1.3

We can construct unitary matrices from either Hermitian or anti-Hermitian matrices by using an exponentiation of the matrix. This formally interprets the exponential series in terms of a matrix $M$ as
\[ \exp(M)\equiv\sum_{n=0}^\infty \frac{M^n}{n!}=I+M+\frac{1}{2}M^2+\frac{1}{6}M^3+\ldots. \]
Given a Hermitian matrix $M$, $M^\dagger=M$, the matrix $U=e^{iM}$ is then automatically unitary since $U^\dagger=e^{-iM^\dagger}=e^{-iM}$ so that $U^\dagger U=e^{-iM}e^{iM}=I$. This is the physics construction that we will mostly use in these notes because of our quantum fondness for Hermitian operators. For an anti-Hermitian matrix $M$, $M^\dagger=-M$ we could instead use that $U=e^M$ is automatically unitary since $U^\dagger=e^{-M}$ so that $U^\dagger U=e^{-M}e^{M}$=I. This is the mathematical construction found in a lot of mathematical literature.

@@

There is a very important class of subgroup called the {\bf normal} subgroup. The importance will become clear in a moment.
\df{A subgroup $H$ is a {\bf normal} (invariant) subgroup, if and only if the {\bf conjugation} of any element by any $g \in G$ is in $H$,\footnote{Another, pretty, but slightly abusive, way of writing the definition of a normal group is to say that $gHg^{-1}=H$. This implies (correctly), that the image if $H$ under the conjugation operation is guaranteed to be the whole of $H$.}
meaning
\[ ghg^{-1} \in H\text{ for all } h \in H.\]
A {\bf simple} group $G$ has no proper normal subgroup. A {\bf semi-simple} group $G$ has no proper abelian normal subgroup.}
We can for example show that for $n>1$, $SU(n)$ is a normal subgroup of $U(n)$, see Ex.~\ref{ex:SUn_normal}.

%%%
\subsection{Quotient groups}
The normal subgroup can be seen as a factor in the original group that can be divided out to form a simpler group that only retains the structure that was not in the normal group. To be more precise we need the concept of {\bf cosets}. 
\df{A {\bf left coset} of a subgroup $H\subset G$ with respect to $g\in G$ is the set of members $\{gh | h\in H\}$, and a {\bf right coset} of the subgroup is the set  $\{hg|h\in H\}$. These are sometimes written $gH$ and $Hg$, respectively.}
For normal subgroups $N$ it can be shown that the sets of left and right cosets $gN$ and $Ng$ coincide and form a group. This is called the {\bf quotient}  or {\bf coset group} and denoted $G/N$.\footnote{Sometimes also called the {\bf factor group}. The notation $G/N$ is pronounced ``G mod N'', where ``mod'' is short for modulo.}
This has as its members all the {\it distinct sets} $\{gn|n\in N\}$, that can be generated by a $g\in G$, and has the binary operation $*$ with $\{gn|n\in N\}*\{g'n|n\in N\}=\{ (g\circ g')n| n\in N\}$. To simplify notation this can be written $gN*g'N=(g\circ g') N$.

To make some sense of this, let us briefly discuss an example of a quotient group. We already know that $SU(n)$ is a normal subgroup to $U(n)$. This means that $U(n)/SU(n)$ is a group. What sort of group is this? Notice that two matrices $U_i$ and $U_j$ live in the same coset of $SU(n)$ if and only if $\det(U_iU_j^{-1})=1$, which is true if and only if $\det(U_i)=\det(U_j)$. In other words, each coset constructed from $SU(n)$ is simply the set of all matrices with a given determinant, which we saw for the $U(n)$ group can be any unit complex number. Observe that this means that many of the cosets are the same set. Here, all cosets generated by members in $U(n)$ with the same determinant is the same coset. 
When these cosets act on each other with the group operation of $U(n)/SU(n)$ they form new cosets of matrices with a determinant that is the product of their individual determinants, {\it i.e.}\ with $U,U'\in U(n)$ we have the product of quotient group member $\{US|S\in SU(n)\}*\{U'S| S\in SU(n)\}=\{ UU'S| S\in SU(n)\}$. Thus, the group behaves exactly as $U(1)$, and is in fact isomorphic to it.

@@

@@

@@

If  we also have $\rho_{12}(g)=0$ in our example we say that the representation is {\bf completely reducible}. It can be shown that in most cases a reducible representation is also completely reducible. In fact, representations for which this is not true tend to be mathematical curiosities. As a result, there is a tendency in physics to use the term ``reducible" where we should use the term ``completely reducible". In the case of a completely reducible representation we can split the vector space $V$ into two vector spaces $V=W_1\oplus W_2$, where $\mathbf v_1\in W_1$ and $\mathbf v_2\in W_2$, and define a representation of $G$ on each of them using $\rho_1$ and $\rho_2$, which in turn could either be reduced more, or would be irreducible.

>>>>

We can construct unitary matrices from either Hermitian or anti-Hermitian matrices by using the {\bf matrix exponential}. This formally interprets the exponential series in terms of a real or complex valued $n\times n$ matrix $M$ as
\[ \exp(M)\equiv\sum_{n=0}^\infty \frac{M^n}{n!}=I+M+\frac{1}{2}M^2+\frac{1}{6}M^3+\ldots. \]
This series can be shown to always converge, so the series is well defined. Since this is a series of non-commuting objects we again have to be careful with using properties of the exponential from ordinary arithmetic. The following useful properties of the matrix exponential can be proven:
\begin{itemize}
\item[i)] $\exp(M^T)=\exp(M)^T$
\item[ii)] $\exp(M^*)=\exp(M)^*$
\item[iii)] If $A$ is an invertible matrix then $\exp(AMA^{-1})=A\exp(M)A^{-1}$
\item[iv)] If $[M,N]=0$ then $\exp(M)\exp(N)=\exp(M+N)$
\end{itemize}

Given a Hermitian matrix $M$, $M^\dagger=M$, the matrix $U=e^{iM}$ is then automatically unitary since $U^\dagger=e^{-iM^\dagger}=e^{-iM}$ so that $U^\dagger U=e^{-iM}e^{iM}=I$, where the last equality is due to $M$ commuting with itself. This is the physics construction that we will mostly use in these notes because of our quantum fondness for Hermitian operators. For an anti-Hermitian matrix $M$, $M^\dagger=-M$, we could instead use that $U=e^M$ is automatically unitary since $U^\dagger=e^{-M}$ so that $U^\dagger U=e^{-M}e^{M}=I$. This is the mathematical construction found in a lot of mathematical literature.

@@

There is a very important class of subgroup called the {\bf normal} subgroup. The importance will become clear in a moment.
\df{A subgroup $H$ is a {\bf normal} (invariant) subgroup, if and only if the {\bf conjugation} of any element $h\in H$ by any $g \in G$ is in $H$,\footnote{Another, pretty, but slightly abusive, way of writing the definition of a normal group is to say that $gHg^{-1}=H$. This implies (correctly), that the image if $H$ under the conjugation operation is guaranteed to be the whole of $H$.}
meaning
\[ ghg^{-1} \in H\text{ for all } h \in H.\]
A {\bf simple} group $G$ has no proper normal subgroup. A {\bf semi-simple} group $G$ has no proper abelian normal subgroup.}
We can for example show that for $n>1$, $SU(n)$ is a normal subgroup of $U(n)$, see Ex.~\ref{ex:SUn_normal}.

%%%
\subsection{Quotient groups}
The normal subgroup can be seen as a factor in the original group that can be divided out to form a simpler group that only retains the structure that was not in the normal group. To be more precise we need the concept of {\bf cosets}. 
\df{A {\bf left coset} of a subgroup $H\subset G$ with respect to $g\in G$ is the set of members $\{gh | h\in H\}$, and a {\bf right coset} of the subgroup is the set  $\{hg|h\in H\}$. These are sometimes written $gH$ and $Hg$, respectively.}
For normal subgroups $H$ it can be shown that the sets of left and right cosets $gH$ and $Hg$ coincide and form a group. This is called the {\bf quotient}  or {\bf coset group} and denoted $G/H$.\footnote{Sometimes also called the {\bf factor group}. The notation $G/H$ is pronounced ``G mod H'', where ``mod'' is short for modulo.}
This has as its members all the {\it distinct sets} $\{gh|h\in H\}$, that can be generated by a $g\in G$, and has the binary operation $*$ with $\{gh|h\in H\}*\{g'h|h\in H\}=\{ (g\circ g')h| h\in H\}$. To simplify notation this can be written $gH*g'H=(g\circ g') H$.

To make some sense of this, let us briefly discuss an example of a quotient group. We already know that $SU(n)$ is a normal subgroup to $U(n)$. This means that $U(n)/SU(n)$ is a group. What sort of group is this? Notice that two matrices $U_i$ and $U_j$ live in the same coset of $SU(n)$ if and only if $\det(U_i)=\det(U_j)$. In other words, each coset constructed from $SU(n)$ is simply the set of all matrices with a given determinant, which we saw for the $U(n)$ group can be any unit complex number. Observe that this means that many of the cosets are the same, all cosets generated by members in $U(n)$ with the same determinant is the same coset. 
When these cosets act on each other with the group operation of $U(n)/SU(n)$ they form new cosets of matrices with a determinant that is the product of their individual determinants, {\it i.e.}\ with $U,U'\in U(n)$ we have the product of quotient group member $\{US|S\in SU(n)\}*\{U'S| S\in SU(n)\}=\{ UU'S| S\in SU(n)\}$. Thus, the group behaves exactly as $U(1)$, and is in fact isomorphic to it.

@@

At the end, let us make a warning: despite the enticing notation it is not in general the case that if $H$ is a normal subgroup to $G$, then $G\cong G/H\times H$.

@@

In physics an important category of representations are the {\bf unitary representations}, which are linear?representations $\rho$ of a group $G$ on a complex Hilbert space $V$ where $\rho(g)$ is a unitary operator, {\it i.e.} an operator that preserves the length of vectors in $V$, for every $g\in G$.

This generalises the unitary (matrix) group, which are examples of unitary representations, beyond finite dimensional vector spaces.

@@

If  we also have $\rho_{12}(g)=0$ in our example we say that the representation is {\bf completely reducible}. It can be shown that in most cases a reducible representation is also completely reducible. In fact, representations for which this is not true tend to be mathematical curiosities. For example, if the vector space is a Hilbert space (states in quantum mechanics), there exist a proof that all representations are completely reducible.  As a result, there is a tendency in physics to use the term ``reducible" where we should use the term ``completely reducible". In the case of a completely reducible representation we can split the vector space $V$ into a direct sum of two vector spaces $V=W_1\oplus W_2$, where $\mathbf v_1\in W_1$ and $\mathbf v_2\in W_2$, and define a representation of $G$ on each of them using $\rho_1$ and $\rho_2$, which in turn could either be reduced more, or would be irreducible.
